Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-24 text/ms.tex
Edit paper, add references, and clean out comments

Black swans are unexpected extreme events with potentially dramatic consequences
\citep{taleb2007,sornette2009}. One of the most striking black swans in ecology
was the asteroid marking the mass extinction of the K-T boundary (REF). Today,
it is the extremeness of climate, in concert with shifts in mean temperature,
that is expected to cause the greatest societal damage \citep{ipcc2012}.
But, while extremes in the physical environment such as wave height, rain
fall, and temperature are present and widely accepted
\citep{gaines1993,katz2005}, it remains unclear the extent to which ecological
systems buffer or suffer from black swans \citep{nunez2012}.

@@

%Considerable work has focussed on heavy tails in physical environmental
%processes in, for example, rain fall (REF), wave height (REF), earth quake
%magnitude, climate temperature (REF). time to extinction, plankton time series,
%GPDD work found little evidence subsequent work suggested, but... Ward applied
%a data-intense method and found evidence for.

%Ecological population dynamic models typically assume 'thin tailed'
%distributions, such as the normal distribution on a multiplicative scale (e.g.
%REF). We define black swan dynamics as cases where jumps in population
%abundance from time step to time step. from heavy-tailed distributions.

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population
dynamics are not heavy tailed (e.g.~REF, REF, REF).
Alternatively, dynamics might be heavy tailed, but a paucity of statistical
methods appropriate for short time series has limited our
ability to detect them.
As four examples: (1) The generalized extreme value (GEV) distribution has been
applied to environmental data (REF).
The GEV describes the distribution of the most extreme
event per time interval (e.g.~heaviest rainfall per year), but this approach
requires long time series that can be condensed into time intervals (REF).
(2) Heavy- and thin-tailed distributions have been competed with each
other (REF), but information criteria likely have
trouble accurately distinguishing such subtle differences in population
dynamics for short time series (REF).
(3) A catastrophic mixture distribution model allows one to quantify the
probability that population events are catastrophic but is also relatively data
intensive \citep{ward2007}.
(4) Other research has focussed on tipping points and regime shifts, but such
methods are usually focussed on entire communities or ecosystems (e.g.\ REF).
Furthermore, black swans are more general than tipping points --- they don't
necessarily imply a permanent change of state.

% For example, a commonly applied probability distribution to detect the shape of
% time series tails (the generalized extreme value distribution) requires lumping
% data into the most extreme event per unit time (e.g.~heaviest rainfall per
% year) and therefore requires a considerable quantity of data (REF). Previous
% large-scale work on the distribution of abundance trends focused on model
% comparison to identify parsimonious descriptions, but has not quantified the
% magnitude of heavy tails on a continuous scale (REF). Furthermore, information
% criteria are poor at distinguishing subtle differences in population dynamics
% for the relatively short time series that are most readily available (REF
% CHANGE THIS?).
%bring in Ward - great method but data intensive, can't be applied to most
%individual time series from the GPDD plankton --- hideous distribution to fit,
%even harder in population dynamics framework -and what about seasonal cycles,
%density dependence? GPDD - model competition and finding most parsimonious
%explanation from set of hypotheses --- but only \emph{very} heavy from not
%heavy GEV --- super data intensive tipping point and regime shift methods ---
%say something\ldots{} not appropriate for most of the data available

@@

% Papers to work in: \citep{inchausti2002, halley2002, inchausti2001}
% \citep{jentsch2007} \citep{ward2007} \citep{garcia-carreras2011}
% \citep{sornette2009} \citep{nunez2012} \citep{thompson2013}
% \citep{beaugrand2012} \citep{pine-iii2009} \citep{doak2008} Extreme rain and
% monsoons over India increasing \citep{goswami2006} \citep{smale2013}
% \citep{easterling2000} \citep{scheffer2003} \citep{katz2005} \citep{taleb2007}
% \citep{vasseur2014} \citep{vert-pre2013} \citep{lindenmayer2010}
% \citep{valpine2002} \citep{gregory2010} \citet{garcia-carreras2011}
% \citeauthor{brook2006} \citetext{\citeyear{brook2006};
%   \citealp{herrandoprez2014}} \citep{sibly2005, ziebarth2010}

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model (REFs?). The Gompertz model represents population growth as a linear
function in $\log$ space. If we let $x_t$ represent the $\log$ abundance at
time $t$, we can represent the Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious model
\citep{brook2006}.

@@

We estimated all models in a Bayesian framework using Stan \citep[][version
2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of Hamiltonian
Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
less correlated chains than algorithms such as the Gibbs sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

@@

% TODO - make sure this is in the supp? However, simultaneously estimating
% observation error
%and process noise in a state space framework is a challenging computational
%problem \citep[e.g.][]{dennis2006, knape2008} even with longer time series and
%without estimating the shape of the process error tails;

@@

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in \overallMinPerc--\overallMaxPerc\%
of populations assessed. We detected black swans more frequently for longer
time series and for populations with a smaller magnitude of process error, as
we might expect. However, we failed to find strong evidence that black swans
were associated with density dependence, population growth rate, or lifespan.
In cases that we verified, black swans were often a result of the interaction
between elements of extreme climate, predators and parasite cycles, and strong
changes in human pressures. Our empirical results, sensitivity tests, and
simulation tests suggest that estimating the tail shape of process error is
a viable method of detecting heavy-tailed population dynamics and if anything
will underestimate the probability of black swans. The presence of black swans
highlights the importance of developing management strategies that detect
quickly, respond to, and are robust to extremes in population dynamics ---
particularly as the frequency and magnitude of climatic extremes increase over
the next century (REF).

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans. A similar focus could benefit resource management once our ability to
predict the spatial probability and covariates of ecological black swans
improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events (REF), and
maintaining genetic, phenotypic, and behavioural diversity may allow some
components of populations persist when others are affected by extreme disease
or environmental forces (REF). Finally, extreme and `unexpected', `surprising', or
`counterintuitive' ecological dynamics offer a tremendous opportunity to learn
about ecological systems, evaluate when our models break down, and tune future
management policy (REF, REF). In recent decades ecology has moved toward
focussing on aspects of variance in addition to mean responses (REF). Our
results and other recent work suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
extreme climate (REF).

%ecological resource management can draw from other disciplines focussed on
%heavy tails, design policy that is robust to heavy tails, and take black swan
%events as opportunities for insight.
%The 'portfolio effect' is often used in the context of dampening variability
%(REFs), but the ultimate portfolio effect is maintaining some stability in the
%face of black swan events.
%unlikely to forecast black swans, at least with any useful advanced warning
%but we might be able to determine probabilities and develop policy that is
%robust to, encourages rapid detection, and that is flexible enough to respond
%quickly to black swans
%example of earth quakes - spatial planning, early warning systems, education
%systems, a general idea about the probability in given places, build buildings
%that are robust
%Suggests that it's important to have policy that is robust to the uncertainty
%about black swans and that can respond and adapt rapidly following black swans
%foot and mouth disease and earthquakes?

>>>>

Black swans are unexpected extreme events with potentially dramatic
consequences \citep{taleb2007,sornette2009}. One of the most striking black
swans in ecology was the asteroid marking the mass extinction of the K-T
boundary \citep{harnik2012}. Today, it is the extremeness of climate, in
concert with shifts in mean temperature, that is expected to cause the greatest
societal damage \citep{ipcc2012}. But, while extremes in the physical
environment such as wave height, rain fall, and temperature are present and
widely accepted \citep{gaines1993,katz2005}, it remains unclear the extent to
which ecological systems buffer or suffer from black swans \citep{nunez2012}.

@@

There are two key reasons why we may find little evidence of ecological black
swans. First, they might not systematically exist. Indeed, the majority of
model fitting and risk forecasting assumes that population dynamics are not
heavy tailed \citep[e.g.][]{brook2006a,dennis2006,knape2012}. Alternatively,
dynamics might be heavy tailed, but a paucity of statistical methods
appropriate for commonly available series has limited our ability to detect
them. For example: (1) The generalized extreme value (GEV) distribution has
been applied to environmental data \citep[e.g.][]{katz2005}. The GEV describes
the distribution of the most extreme event per time interval (e.g.~heaviest
rainfall per year), but this approach requires long time series that can be
condensed into time intervals. (2) Thin- and heavy-tailed distributions have
been competed with each other in an information theoretic framework
\citep{halley2002}, but this analysis does not quantify the probability of
black swans on a continuous scale or allow for population dynamics. (3) Fitting
a catastrophic mixture distribution state space model allows one to quantify
the probability that population events are catastrophic but is also data
intensive \citep{ward2007}.

@@

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance at time $t$, we can represent the Gompertz
model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious population model
\citep{brook2006}.

@@

We estimated all models in a Bayesian framework using Stan \citep[][version
2.4.0]{stan-manual2014} via the R computing environment \citep{r2014}. Stan
samples from the posterior distribution with an adaptive version of Hamiltonian
Markov chain Monte Carlo called the No-U-Turn Sampler and generally obtains
less correlated samples than algorithms such as the Gibbs sampler
\citep{hoffman2014}. We assured that chains had sufficiently converged and the
sampler had obtained sufficient independent samples from the posterior
($\widehat{R} < 1.05$, $n_\mathrm{eff} > 200$; Supporting Material). We used
simulated data to test how easily we could detect $\nu$ given different sample
sizes and to ensure we could recover unbiased parameter estimates from the
Gompertz model (Supporting Material, Fig.~\ref{fig:sim-nu},
\ref{fig:sim-gompertz}, \ref{fig:sim-gompertz-boxplots}).

@@

@@

We found strong evidence that the multiplicative process noise jumps from time
step to time step were greater than we would expect under a normal distribution
and a Gompertz or Ricker-logistic model in \overallMinPerc--\overallMaxPerc\%
of populations assessed. We detected black swans more frequently for longer
time series and for populations with a smaller magnitude of process error, as
we might expect. However, we failed to find strong evidence that black swans
were associated with density dependence, population growth rate, or lifespan.
In cases that we verified, black swans were often a result of the interaction
between elements of extreme climate, predators and parasite cycles, and strong
changes in human pressures. Our empirical results, sensitivity tests, and
simulation tests suggest that estimating the tail shape of process error is
a viable method of detecting heavy-tailed population dynamics and if anything
will underestimate the probability of black swans. The presence of black swans
highlights the importance of developing management strategies that detect
quickly, respond to, and are robust to extremes in population dynamics ---
particularly as the frequency and magnitude of climatic extremes increase over
the next century \citep{easterling2000,ipcc2012}.

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans. A similar focus could benefit resource management once our ability to
predict the spatial probability and covariates of ecological black swans
improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by extreme disease or environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and `unexpected',
`surprising', or `counterintuitive' ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and tune future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
extreme climate \citep{meehl2004,ipcc2012}.
